---
title: "Graphics"
author: "DJM"
date: "11 April 2017"
output:
  slidy_presentation: default
  pdf_document: default
  beamer_presentation: default
---





## Making better graphics


```{r setup, message=FALSE, echo=FALSE}
# Need the knitr package to set chunk options
library(knitr)
library(tidyverse)
# Set knitr options for knitting code into the report:
# - Don't print out code (echo)
# - Save results so that code blocks aren't re-run unless code changes (cache),
# _or_ a relevant earlier code block changed (autodep), but don't re-run if the
# only thing that changed was the comments (cache.comments)
# - Don't clutter R output with messages or warnings (message, warning)
  # This _will_ leave error messages showing up in the knitted report
opts_chunk$set(message=FALSE, warning=FALSE, fig.align='center',fig.width=8,
               fig.height=4,cache=TRUE,autodep=TRUE)
```

> "The simple graph has brought more information to the data analyst’s mind 
> than any other device." --- John Tukey

* This week we're going to learn how to use `ggplot2`, Hadley Wickham's package for making graphics.

* This material is based on Chapter 3 of his book "`R` for data science".

* That is your reading for this week (link on syllabus page).


## Why ggplot?

* `ggplot2` is a bit hard to get used to.

* But it's figures are much better looking than `R`'s `plot` functions.

* Compare:

```{r plot-ex-data, echo=FALSE}
nums = rbinom(250,2,.66)+1
classes = LETTERS[nums]
vals = rnorm(250, nums, 1/nums)
df = data.frame(classes, vals)
```

```{r plots}
boxplot(vals~classes, data=df)
ggplot(df,mapping=aes(x=classes,y=vals,fill=classes)) + geom_boxplot()
```

## Graphics are for addressing questions

* Do cars with big engines use more fuel than cars with small engines? 

* You probably already have an answer, but try to make your answer precise. 

* What does the relationship between engine size and fuel efficiency look like? 

* Is it positive? Negative? Linear? Nonlinear?


## Some data

```{r}
mpg
```

Among the variables in `mpg` are:

1. `displ`, a car's engine size, in litres.

1. `hwy`, a car's fuel efficiency on the highway, in miles per gallon (mpg). 
  A car with a low fuel efficiency consumes more fuel than a car with a high 
  fuel efficiency when they travel the same distance. 
  
## First plot

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

* `ggplot()` creates a coordinate system. 

* The first argument of `ggplot()` is the dataset to use in the graph. So `ggplot(data = mpg)` creates an empty graph, but it's not very interesting.

* `geom_point()` adds a layer of points to your plot.

* Each geom function in ggplot2 takes a `mapping` argument. 

* This defines how variables in your dataset are mapped to visual properties. 

* The `mapping` argument is always paired with `aes()`, and the `x` and `y` arguments of `aes()` specify which variables to map to the x and y axes. 

* ggplot2 looks for the mapped variable in the `data` argument, in this case, `mpg`.

## Again...

Note:

```{r}
ggplot(mpg) + geom_point(aes(displ,hwy))
```

In general:

```
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

## Aesthetics

> "The greatest value of a picture is when it forces us to notice what we
> never expected to see." --- John Tukey

```{r, echo = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() + 
  geom_point(data = dplyr::filter(mpg, displ > 5, hwy > 20), colour = "red", size = 2.2)
```

* The red points seem to be outliers.

## Coloring

* Perhaps another variable explains them (say, hybrid SUVs)

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```


## Size

```{r, warning=TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

## Transparency and shape

```{r out.width = "50%", fig.align = 'default', warning = FALSE, fig.asp = 1/2, fig.cap ="",fig.show='hold',warning=TRUE}
# Left
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

# Right
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

## Just like some color

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "#31B404")
```

## Facets

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```


```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

## Rows or columns

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)
    
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

## What are geoms?

```{r, out.width = "50%", fig.align="default", fig.show='hold', message = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

* Different __geoms__ lead to different kinds of geometric objects.

* bar charts use bar geoms, 

* line charts use line geoms, 

* boxplots use boxplot geoms

* Scatterplots break the trend; they use the point geom.

## Line types

```{r message = FALSE}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

* One line describes all of the points with a `4` value, 

* one line describes all of the points with an `f` value, 

* and one line describes all of the points with an `r` value.

## More clear

```{r, echo=FALSE, message = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() +
  geom_smooth(mapping = aes(linetype = drv))
```


## Grouping

* Many geoms use a single geometric object to display multiple rows of data. 

* you can set the `group` aesthetic to a categorical variable to draw multiple objects. 

* ggplot2 will draw a separate object for each unique value of the grouping variable. 

* the `group` aesthetic by itself does not add a legend or distinguishing features to the geoms (`linetype` or `color` do).

```{r, out.width = "50%", fig.show='hold',message = FALSE}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
              
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```

## Multiple geoms

* To display multiple geoms in the same plot, add multiple geom functions to `ggplot()`:

```{r, message = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

* Creates duplication, annoying to type `aes(x = displ, y = hwy)`.

```{r, eval = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```

* mappings in `ggplot` are _global_, in a geom, they are _local_ to the geom.

## Ex

```{r, message = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

```{r, message = FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

## Bar charts

* The `diamonds` dataset comes in ggplot2 and contains information about ~54,000 diamonds, 

* Information about the `price`, `carat`, `color`, `clarity`, and `cut` of each diamond.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

Some graphs, like bar charts, calculate new values to plot:

* bar charts, histograms, and frequency polygons bin your data 
  and then plot bin counts, the number of points that fall in each bin.

* smoothers fit a model to your data and then plot predictions from the
  model.

* boxplots compute a robust summary of the distribution and then display a 
  specially formatted box.

The algorithm used to calculate new values for a graph is called a __stat__, short for statistical transformation.

## Some stats

```{r}
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

* Every geom has a default stat.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))
```

```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
    )
```

* ggplot2 provides over 20 stats for you to use.

## Looking better

```{r message = FALSE,fig.show='hold'}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```

## Stacking

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```

* The stacking is performed automatically by the __position adjustment__ specified by the `position` argument. 

* If you don't want a stacked bar chart, you can use one of three other options: `"identity"`, `"dodge"` or `"fill"`.

## Identity

* `position = "identity"` will place each object exactly where it falls in 
    the context of the graph. 
    
* This is not very useful for bars, because it
overlaps them. 

* To see that overlapping we either need to make the bars
slightly transparent by setting `alpha` to a small value, or completely
transparent by setting `fill = NA`.
    
```{r message = FALSE,fig.show='hold'}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```


* The identity position adjustment is more useful for 2d geoms, like points, where it is the default.

## Fill

*   `position = "fill"` works like stacking, but makes each set of stacked bars
the same height. 
    
* This makes it easier to compare proportions acrossgroups.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

## Dodge

*   `position = "dodge"` places overlapping objects directly _beside_ one another. 
    
* This makes it easier to compare individual values.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

## Jitter

```{r echo = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

* the plot displays only 126 points, even though there are 234 observations in the dataset

* The values of `hwy` and `displ` are rounded so the points appear on a grid and many points overlap each other. 

* This problem is known as __overplotting__. 

* This arrangement makes it hard to see where the mass of the data is. 

* `position = "jitter"` adds a small amount of random noise to each point. 


```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```

## Coordinate systems

* Coordinate systems are probably the most complicated part of ggplot2. 

* The default coordinate system is the Cartesian coordinate system where the x and y position act independently to find the location of each point. 

* There are a number of other coordinate systems that are occasionally helpful.

*   `coord_flip()` switches the x and y axes. 

* This is useful (for example), if you want horizontal boxplots. 

* It's also useful for long labels: it's hard to get them to fit without overlapping on the x-axis.
    
```{r message = FALSE,fig.show='hold'}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot()
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot() + coord_flip()
```

## Quickmap

* `coord_quickmap()` sets the aspect ratio correctly for maps. 

* important if you're plotting spatial data 

```{r message = FALSE,fig.show='hold'}
nz <- map_data("nz")
ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black")
ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_quickmap()
```

## Polar

* `coord_polar()` uses polar coordinates. 

* Polar coordinates reveal an 
    interesting connection between a bar chart and a Coxcomb chart.
    
* I find this figure stupid, but some people like it.
    
```{r fig.width = 3, out.width = "50%", fig.align = "default", fig.asp = 1}
## ggplot actually returns "invisibly"
bar <- ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut), 
    show.legend = FALSE, width = 1) + 
    theme(aspect.ratio = 1) +
    labs(x = NULL, y = NULL)

## Note this odd syntax    
bar + coord_flip()
bar + coord_polar()
```

## Summary

```
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```

## Another example

```{r include=FALSE}
db1 = read.csv('~/Documents/Work/Teaching and TA/s432/projectData/IUFootball/defensiveback1.csv', 
               stringsAsFactors=FALSE) # path to the data (original)
library(lubridate)
db1$Start.DateTime = mdy_hm(db1$Start.DateTime)
db1$weekday = weekdays(db1$Start.DateTime)
db1$BeforeIowa = db1$Start.DateTime <= dmy('07/11/2015')
```


```{r}
boxplot(Mechanical.Load~weekday+BeforeIowa, data=db1) # ugly, hard to read
```

## ggplot

```{r}
ggplot(db1, mapping=aes(x=weekday, y=Mechanical.Load)) +
  geom_boxplot(aes(fill=BeforeIowa))
```

## Make weekdays in the correct order

I googled `R reorder factor levels`

```{r}
db1$weekday = factor(db1$weekday, c('Monday','Tuesday','Wednesday','Thursday','Friday','Saturday','Sunday'))
ggplot(db1, mapping=aes(x=weekday, y=Mechanical.Load)) +
  geom_boxplot(aes(fill=BeforeIowa))
```

## these colors are ugly (log scale?)

I googled `R ggplot2 fill colors`

```{r}
ggplot(db1, mapping=aes(x=weekday, y=log(Mechanical.Load))) + geom_boxplot(aes(fill=BeforeIowa)) + 
  scale_fill_brewer()
```

## still ugly, log is dumb

Try [`?scale_fill_brewer`](http://docs.ggplot2.org/current/scale_brewer.html) to see options, pretty opaque except for the examples

```{r}
ggplot(db1, mapping=aes(x=weekday, y=Mechanical.Load)) + geom_boxplot(aes(fill=BeforeIowa)) +
  scale_fill_brewer(palette='Set1')
```

## Not so bad

What if I made `weekday` the colors and `BeforeIowa` on the `x`-axis?

```{r}
ggplot(db1, mapping=aes(x=BeforeIowa, y=Mechanical.Load)) + geom_boxplot(aes(fill=weekday)) +
  scale_fill_brewer(palette='Set1')
```

## Stick with the previous, rename labels

```{r}
db1$BeforeIowa = factor(db1$BeforeIowa, labels=c('After Iowa','Before Iowa'))
ggplot(db1, mapping=aes(x=weekday, y=Mechanical.Load)) + geom_boxplot(aes(fill=BeforeIowa)) +
  scale_fill_brewer(palette='Set1') + 
  labs(y='Mechanical Load', x=element_blank(), fill=element_blank()) + 
  theme(legend.position='top')
```

## All the players in one dataset

```{r, include=FALSE}
projDir = '~/Documents/Work/Teaching and TA/s432/projectData/IUFootball/'
files = list.files(projDir, pattern='.csv') # finds .csv files in directory 'projDir'
dnames = unlist(strsplit(files, '.csv')) # gets only the 'handle'
for(i in 1:length(dnames)) assign(dnames[i], read.csv(paste0(projDir, files[i]), stringsAsFactors = FALSE))
library(dplyr)
workout = bind_rows(defensiveback1, linebacker1, linebacker2, widereceiver1)
grades = bind_rows(defensiveback1grades, linebacker1grades, linebacker2grades, widereceiver1grades, .id='Subject.Name')
workout$Start.DateTime = mdy_hm(workout$Start.DateTime)
workout$weekday = weekdays(workout$Start.DateTime)
workout$BeforeIowa = workout$Start.DateTime <=  dmy('07/11/2015')
workout$BeforeIowa = factor(workout$BeforeIowa, labels=c('After Iowa','Before Iowa'))
workout$weekday = factor(workout$weekday, c('Monday','Tuesday','Wednesday','Thursday','Friday','Saturday','Sunday'))
```


```{r,fig.height=15}
ggplot(workout, mapping=aes(x=weekday, y=Mechanical.Load)) + geom_boxplot(aes(fill=BeforeIowa)) +
  scale_fill_brewer(palette='Set1') + 
  labs(y='Mechanical Load', x=element_blank(), fill=element_blank()) + 
  theme(legend.position='top') + facet_wrap(~Subject.Name, nrow=2)
```
